---
title: "Getting Started with SurplusProductionModelMSE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with SurplusProductionModelMSE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

The **SurplusProductionModelMSE** package provides a closed-loop Management Strategy Evaluation
framework. It uses the
**SurplusProductionModel** package as the assessment engine and supports
spatial structure mismatch testing between operating and estimation models.

A Management Strategy Evaluation simulates the full feedback loop of fishery
management: the *operating model* (OM) represents the "true" population, a
*harvest control rule* (HCR) sets catch limits, *observation error* adds
realistic noise to monitoring data, and an optional *estimation model* (EM)
re-estimates stock status each year. By running many stochastic replicates,
MSE quantifies expected trade-offs among management objectives—yield,
conservation risk, and catch stability.

Key capabilities:

- Forward projection with Pella-Tomlinson population dynamics (single- or multi-area)
- Four built-in harvest control rules (constant F, constant catch, hockey stick,
  CCAMLR krill-style)
- Observation error calibrated from fitted model residuals
- Implementation error with autocorrelation and maximum overage caps
- Performance metrics: depletion risk, yield, biomass ratios, catch variability
- Spatial robustness testing with OM/EM structural mismatch
- Trade-off visualisation with Pareto frontiers

## Installation

```r
# Install from GitHub (requires SurplusProductionModel)
# install.packages("devtools")
devtools::install_github("alistairdunn1/SurplusProductionModel")
devtools::install_github("alistairdunn1/SurplusProductionModelMSE")
```

## Quick Start: A Single-Area MSE

This guide walks through a complete MSE from configuration to results
interpretation, using a simple single-area Schaefer model.

### Step 1 — Define the Operating Model

The operating model specifies the "true" population. We use parameters
loosely based on Ross Sea Antarctic toothfish:

```{r define-om}
library(SurplusProductionModelMSE)

om <- om_config(
  n_areas = 1L,
  true_params = list(
    r = 0.3,         # intrinsic growth rate
    K = 5000,        # carrying capacity (tonnes)
    m = 2,           # shape (m=2 is the Schaefer model)
    sigma_obs = 0.2, # observation error SD (log-scale)
    q = 0.001,       # CPUE catchability
    B0 = 5000        # initial biomass = K (unfished)
  )
)

print(om)
```

With the Schaefer model ($m = 2$), biological reference points are:

$$
B_{\text{MSY}} = \frac{K}{2}, \quad
F_{\text{MSY}} = \frac{r}{2m} = \frac{r}{4}, \quad
\text{MSY} = \frac{r K}{8}
$$

For our parameters: $B_{\text{MSY}}$ = 2500 t, $F_{\text{MSY}}$ = 0.075,
MSY = 187.5 t.

### Step 2 — Choose Harvest Control Rules

We compare two management strategies: a constant catch rule and a
hockey-stick rule that reduces catch when biomass declines.

```{r define-hcrs}
# Strategy 1: constant catch at 80% of MSY
hcr_const <- hcr_constant_catch(catch_target = 150)

# Strategy 2: hockey-stick that ramps F down when B < 0.4K
hcr_hs <- hcr_hockey_stick(
  f_target = 0.075,   # target F = FMSY
  b_limit  = 0.20,    # F = 0 below 20% K
  b_target = 0.40     # full F above 40% K
)
```

### Step 3 — Create Scenarios

Each scenario combines an HCR with optional implementation error. Here we
add moderate implementation error (CV = 10%, no bias):

```{r define-scenarios}
ie <- impl_error(cv = 0.1, bias = 1.0, autocorr = 0, max_overage = 1.2)

scenarios <- list(
  create_scenario(
    name = "Constant catch",
    harvest_control_rule = hcr_const,
    implementation_error = ie
  ),
  create_scenario(
    name = "Hockey stick",
    harvest_control_rule = hcr_hs,
    implementation_error = ie
  )
)
```

### Step 4 — Run the MSE

Run 50 stochastic replicates over a 20-year projection horizon. We skip the
estimation model in this example to focus on the feedback loop mechanics:

```{r run-mse}
set.seed(42)

result <- mse_simulation(
  operating_model  = om,
  scenarios        = scenarios,
  n_sims           = 50L,
  n_proj_years     = 20L,
  min_assess_years = 99L,   # skip EM fitting (use true biomass)
  seed             = 42L,
  verbose          = FALSE
)

print(result)
```

Setting `min_assess_years = 99L` means the estimation model is never invoked
and the HCR receives the true biomass. This is useful for understanding HCR
behaviour in isolation, without the additional noise from stock assessment
estimation error.

### Step 5 — Evaluate Performance

Calculate performance metrics for each scenario. The default thresholds
flag the probability of biomass falling below 50% and 20% of $K$:

```{r performance}
perf_const <- calculate_performance_metrics(
  trajectories = result$results[["Constant catch"]]$trajectories,
  om_config    = om,
  thresholds   = c(0.5, 0.2)
)

perf_hs <- calculate_performance_metrics(
  trajectories = result$results[["Hockey stick"]]$trajectories,
  om_config    = om,
  thresholds   = c(0.5, 0.2)
)

cat("=== Constant Catch ===\n")
summary(perf_const)

cat("\n=== Hockey Stick ===\n")
summary(perf_hs)
```

### Step 6 — Visualise Trade-Offs

The `compare_scenarios()` function creates a trade-off plot with Pareto
frontier identification. First, attach the performance objects to the MSE
result:

```{r tradeoff, fig.height = 5}
# Attach performance to result for compare_scenarios
result$results[["Constant catch"]]$performance <- perf_const
result$results[["Hockey stick"]]$performance <- perf_hs

p <- compare_scenarios(
  mse_results = result,
  metric_x    = "mean_catch",
  metric_y    = "Pr(B<20%B0)_ever"
)
print(p)
```

Points on the Pareto frontier represent strategies where no alternative
achieves higher catch without also increasing risk (or vice versa).

## Biomass Trajectory Plots

We can extract and plot the biomass trajectories from the MSE result to
visualise population dynamics under each strategy:

```{r trajectory-plot, fig.height = 4}
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))

for (sc_name in c("Constant catch", "Hockey stick")) {
  bio_raw <- result$results[[sc_name]]$trajectories$biomass
  # biomass may be 3D [sims x years x areas]; collapse to 2D for single-area
  if (length(dim(bio_raw)) == 3) {
    bio <- bio_raw[, , 1]
  } else {
    bio <- bio_raw
  }
  n_sims <- nrow(bio)
  n_yrs  <- ncol(bio)
  years  <- seq_len(n_yrs)

  plot(NA, xlim = c(1, n_yrs), ylim = c(0, max(bio, na.rm = TRUE)),
       xlab = "Projection Year", ylab = "Biomass (t)", main = sc_name)
  for (i in seq_len(min(n_sims, 20))) {
    lines(years, bio[i, ], col = adjustcolor("steelblue", 0.3))
  }
  # Median trajectory
  med <- apply(bio, 2, median, na.rm = TRUE)
  lines(years, med, col = "darkblue", lwd = 2)
  abline(h = 5000 * 0.5, lty = 2, col = "red")  # BMSY
  abline(h = 5000 * 0.2, lty = 3, col = "red")   # 20% K limit
}
```

## Harvest Control Rules in Detail

The MSE package provides four HCR factory functions, each returning a
closure with signature `function(biomass, reference_points)`:
 
| HCR | Description |
|-----|-------------|
| `hcr_constant_f(f_target)` | Apply a fixed fishing mortality rate |
| `hcr_constant_catch(catch_target)` | Apply a fixed catch regardless of biomass |
| `hcr_hockey_stick(f_target, b_limit, b_target)` | Ramp F linearly between limit and target depletion |
| `hcr_ccamlr_krill(gamma, target_depletion, limit_depletion)` | CCAMLR-style precautionary rule |

The hockey-stick rule is particularly useful for CCAMLR-style management:

$$
F =
\begin{cases}
0 & \text{if } B/K \leq b_{\text{limit}} \\
F_{\text{target}} \cdot \frac{B/K - b_{\text{limit}}}{b_{\text{target}} - b_{\text{limit}}} & \text{if } b_{\text{limit}} < B/K < b_{\text{target}} \\
F_{\text{target}} & \text{if } B/K \geq b_{\text{target}}
\end{cases}
$$

## Multi-Area Spatial Models

The framework supports spatial structure in the operating model. Areas are
connected via a movement kernel built from a distance matrix:

```{r spatial-om}
om_spatial <- om_config(
  n_areas         = 2L,
  movement_rate   = 0.1,
  distance_matrix = matrix(c(0, 100, 100, 0), nrow = 2),
  true_params     = list(
    r = 0.3, K = 5000, m = 2,
    sigma_obs = 0.2, q = 0.001, B0 = 5000
  )
)
```

The movement kernel is computed from the distance matrix:

```{r movement-kernel}
kernel <- build_movement_kernel(
  distance_matrix = matrix(c(0, 100, 100, 0), nrow = 2),
  decay = 0.01
)
print(kernel)
```

### Spatial Robustness: Self-Test

A *self-test* verifies that when the EM has the same spatial structure as the
OM, the management procedure performs as expected. This is a sanity check
before testing spatial misspecification:

```{r self-test, eval = FALSE}
st <- run_self_test(
  operating_model      = om,
  harvest_control_rule = hcr_const,
  n_sims               = 20L,
  n_proj_years         = 20L,
  seed                 = 123L
)
print(st)
```

If `biomass_conserved` and `positive_biomass` are both `TRUE`, the model
machinery is working correctly.

## Implementation Error

The `impl_error()` specification adds realism by modelling the difference
between the intended TAC and realised catch:

```{r impl-error-example}
ie_ar1 <- impl_error(
  cv          = 0.15,     # coefficient of variation
  bias        = 1.0,      # no systematic bias
  autocorr    = 0.3,      # year-to-year correlation
  max_overage = 1.2       # cap realised catch at 120% of TAC
)

# Apply to a single TAC
set.seed(1)
result_ie <- apply_implementation_error(
  tac          = 200,
  impl_config  = ie_ar1,
  previous_eps = 0,
  seed         = 1L
)
cat(sprintf("TAC = 200, Realised catch = %.1f\n", result_ie$catch))
```

## Observation Error

Observation error parameters can be calibrated from a previously fitted
model, or specified directly:

```{r obs-error-manual}
obs_params <- structure(
  list(
    sigma  = 0.2,
    rho    = 0,
    labels = NULL,
    n_obs  = 20L
  ),
  class = "obs_error_params"
)
print(obs_params)
```

When a fitted `ProductionModel` is available, use `calibrate_observation_error()`
to extract residual-based parameters automatically.

## Model Formulation

The operating model uses the Pella-Tomlinson surplus production model:

**Production function:**

$$P(B) = \frac{r \, B}{m} \left[ 1 - \left(\frac{B}{K}\right)^{m-1} \right]$$

**State equation:**

$$B_{t+1} = B_t + P(B_t) - C_t$$

**For multi-area models**, biomass is redistributed according to a movement
kernel $\mathbf{M}$ at each time step:

$$\mathbf{B}_{t+1} = \mathbf{M} \cdot \left[ \mathbf{B}_t + P(\mathbf{B}_t) - \mathbf{C}_t \right]$$

**Reference points** (Pella-Tomlinson):

$$
B_{\text{MSY}} = K \left(\frac{1}{m}\right)^{1/(m-1)}, \quad
F_{\text{MSY}} = \frac{r(1 - 1/m)}{m}, \quad
\text{MSY} = F_{\text{MSY}} \times B_{\text{MSY}}
$$

## Next Steps

- Add an estimation model via `em_config()` to test assessment-based feedback
- Explore spatial misspecification (e.g. 2-area OM with 1-area EM)
- Run `compare_scenarios()` to evaluate Pareto-optimal management strategies
- Use `equilibrium_f()` to compute equilibrium fishing mortality at any
  depletion level

```{r session-info}
sessionInfo()
```
