---
title: "Getting Started with SurplusProductionModelMSE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with SurplusProductionModelMSE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

The **SurplusProductionModelMSE** package provides a closed-loop Management Strategy Evaluation
framework. It uses the
**SurplusProductionModel** package as the assessment engine and supports
spatial structure mismatch testing between operating and estimation models.

A Management Strategy Evaluation simulates the full feedback loop of fishery
management: the *operating model* (OM) represents the "true" population, a
*harvest control rule* (HCR) sets catch limits, *observation error* adds
realistic noise to monitoring data, and an optional *estimation model* (EM)
re-estimates stock status each year. By running many stochastic replicates,
MSE quantifies expected trade-offs among management objectives—yield,
conservation risk, and catch stability.

Key capabilities:

- Forward projection with Pella-Tomlinson population dynamics (single- or multi-area)
- Four built-in harvest control rules (constant F, constant catch, hockey stick,
  CCAMLR krill-style)
- Observation error calibrated from fitted model residuals
- Implementation error with autocorrelation and maximum overage caps
- Performance metrics: depletion risk, yield, biomass ratios, catch variability
- Spatial robustness testing with OM/EM structural mismatch
- Trade-off visualisation with Pareto frontiers

## Installation

```r
# Install from GitHub (requires SurplusProductionModel)
# install.packages("devtools")
devtools::install_github("alistairdunn1/SurplusProductionModel")
devtools::install_github("alistairdunn1/SurplusProductionModelMSE")
```

## Quick Start: A Single-Area MSE

This guide walks through a complete MSE from configuration to results
interpretation, using a simple single-area Schaefer model.

### Step 1 — Define the Operating Model

The operating model specifies the "true" population. We use parameters
loosely based on Ross Sea Antarctic toothfish:

```{r define-om}
library(SurplusProductionModelMSE)

om <- om_config(
  n_areas = 1L,
  true_params = list(
    r = 0.3,         # intrinsic growth rate
    K = 5000,        # carrying capacity (tonnes)
    m = 2,           # shape (m=2 is the Schaefer model)
    sigma_obs = 0.2, # observation error SD (log-scale)
    q = 0.001,       # CPUE catchability
    B_initial = 5000 # initial biomass = K (unfished, d0 = 1)
  )
)

print(om)
```

With the standard Schaefer model ($m = 2$), biological reference points are:

$$
B_{\text{MSY}} = \frac{K}{2}, \quad
F_{\text{MSY}} = \frac{r}{m} = \frac{r}{2}, \quad
\text{MSY} = \frac{r K}{4}
$$

For our parameters: $B_{\text{MSY}}$ = 2500 t, $F_{\text{MSY}}$ = 0.15,
MSY = 375 t.

### Step 2 — Build a Historical Data Series

Surplus production models are only weakly identifiable from an index that
shows little exploitation contrast: fishing the stock near a constant,
sustainable catch leaves CPUE almost flat, and the estimation model (EM)
cannot separate $r$ from $K$. In practice an assessment is conditioned on an
observed history that shows the stock being fished down from an earlier,
less-exploited state. We construct that history by projecting the operating
model under a historical catch series with `project_trajectory()`, then
generating an observed CPUE series from the resulting biomass with
`simulate_cpue()`:

```{r historical-data}
set.seed(11)

# Ramp catch up toward and beyond MSY over ten years, then hold near MSY;
# this depletes the stock from its unfished state and gives the EM genuine
# contrast from which to estimate r and K.
historical_catch <- c(seq(100, 650, length.out = 10), rep(375, 5))
historical_years <- seq_along(historical_catch)

historical_biomass <- project_trajectory(
  B_initial    = om$true_params$B_initial,
  catch_series = historical_catch,
  om_config    = om,
  seed         = 11
)

historical_cpue <- simulate_cpue(
  true_biomass     = historical_biomass[historical_years, 1],
  obs_error_params = list(sigma = om$true_params$sigma_obs, rho = 0),
  q                = om$true_params$q,
  years            = historical_years,
  seed             = 21
)

historical_data <- list(
  catch_data = data.frame(year = historical_years, catch = historical_catch),
  cpue_data  = data.frame(year = historical_cpue$year, cpue = historical_cpue$cpue)
)

# The projection phase begins where the historical period ends, not at the
# unfished state, so the operating model's starting biomass is updated to match.
om$true_params$B_initial <- tail(historical_biomass[, 1], 1)

cat("Depletion at start of projection:",
    round(om$true_params$B_initial / om$true_params$K, 2), "K\n")
```

### Step 3 — Choose Harvest Control Rules

We compare two management strategies: a constant catch rule and a
hockey-stick rule that reduces catch when biomass declines.

```{r define-hcrs}
# Strategy 1: constant catch at 80% of MSY (0.8 * 375 = 300)
hcr_const <- hcr_constant_catch(catch_target = 300)

# Strategy 2: hockey-stick that ramps F down when B < 0.4K
hcr_hs <- hcr_hockey_stick(
  f_target = 0.15,    # target F = FMSY = r/m
  b_limit  = 0.20,    # F = 0 below 20% K
  b_target = 0.40     # full F above 40% K
)
```

### Step 4 — Create Scenarios

Each scenario combines an HCR with optional implementation error. Here we
add moderate implementation error (CV = 10%, no bias):

```{r define-scenarios}
ie <- impl_error(cv = 0.1, bias = 1.0, autocorr = 0, max_overage = 1.2)

scenarios <- list(
  create_scenario(
    name = "Constant catch",
    harvest_control_rule = hcr_const,
    implementation_error = ie
  ),
  create_scenario(
    name = "Hockey stick",
    harvest_control_rule = hcr_hs,
    implementation_error = ie
  )
)
```

### Step 5 — Run the MSE

Run 50 stochastic replicates over a 20-year projection horizon, conditioned on
the historical data series from Step 2. An assessment is run each year so
that the harvest control rule responds to the estimated stock state:

```{r run-mse}
set.seed(42)

result <- mse_simulation(
  operating_model  = om,
  scenarios        = scenarios,
  n_sims           = 50L,
  n_proj_years     = 20L,
  historical_data  = historical_data,
  min_assess_years = 3L,    # safety floor; already satisfied by historical_data
  initial_tac      = 300,   # catch limit before the first assessment (~0.8 * MSY)
  seed             = 42L,
  verbose          = FALSE
)

print(result)
```

Supplying `historical_data` triggers an initial assessment before the
projection begins, fitted against the historical catch and CPUE series from
Step 2. `min_assess_years` sets a floor on how much accumulated data is
needed before any assessment can be fitted, but the historical series already
exceeds it, so an assessment is performed every subsequent year of the
projection and the harvest control rule responds to the current estimated
biomass. `initial_tac` is unused here because the historical assessment
already supplies a first estimate; it would apply only if `historical_data`
were omitted. Because no `estimation_model` is supplied, a default surplus
production estimation model is fitted at each assessment.

### Step 6 — Evaluate Performance

Calculate performance metrics for each scenario. The default thresholds
flag the probability of biomass falling below 50% and 20% of $K$:

```{r performance}
perf_const <- calculate_performance_metrics(
  trajectories = result$results[["Constant catch"]]$trajectories,
  om_config    = om,
  thresholds   = c(0.5, 0.2)
)

perf_hs <- calculate_performance_metrics(
  trajectories = result$results[["Hockey stick"]]$trajectories,
  om_config    = om,
  thresholds   = c(0.5, 0.2)
)

cat("=== Constant Catch ===\n")
summary(perf_const)

cat("\n=== Hockey Stick ===\n")
summary(perf_hs)
```

### Step 7 — Visualise Trade-Offs

The `compare_scenarios()` function creates a trade-off plot with Pareto
frontier identification. First, attach the performance objects to the MSE
result:

```{r tradeoff, fig.height = 5}
# Attach performance to result for compare_scenarios
result$results[["Constant catch"]]$performance <- perf_const
result$results[["Hockey stick"]]$performance <- perf_hs

p <- compare_scenarios(
  mse_results = result,
  metric_x    = "mean_catch",
  metric_y    = "Pr(B<20%K)_ever"
)
print(p)
```

Points on the Pareto frontier represent strategies where no alternative
achieves higher catch without also increasing risk (or vice versa).

## Biomass Trajectory Plots

We can extract and plot the biomass trajectories from the MSE result to
visualise population dynamics under each strategy:

```{r trajectory-plot, fig.height = 4}
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))

for (sc_name in c("Constant catch", "Hockey stick")) {
  bio_raw <- result$results[[sc_name]]$trajectories$biomass
  # biomass may be 3D [sims x years x areas]; collapse to 2D for single-area
  if (length(dim(bio_raw)) == 3) {
    bio <- bio_raw[, , 1]
  } else {
    bio <- bio_raw
  }
  n_sims <- nrow(bio)
  n_yrs  <- ncol(bio)
  years  <- seq_len(n_yrs)

  plot(NA, xlim = c(1, n_yrs), ylim = c(0, max(bio, na.rm = TRUE)),
       xlab = "Projection Year", ylab = "Biomass (t)", main = sc_name)
  for (i in seq_len(min(n_sims, 20))) {
    lines(years, bio[i, ], col = adjustcolor("steelblue", 0.3))
  }
  # Median trajectory
  med <- apply(bio, 2, median, na.rm = TRUE)
  lines(years, med, col = "darkblue", lwd = 2)
  abline(h = 5000 * 0.5, lty = 2, col = "red")  # BMSY
  abline(h = 5000 * 0.2, lty = 3, col = "red")   # 20% K limit
}
```

## Harvest Control Rules in Detail

The MSE package provides four HCR factory functions, each returning a
closure with signature `function(biomass, reference_points)`:
 
| HCR | Description |
|-----|-------------|
| `hcr_constant_f(f_target)` | Apply a fixed fishing mortality rate |
| `hcr_constant_catch(catch_target)` | Apply a fixed catch regardless of biomass |
| `hcr_hockey_stick(f_target, b_limit, b_target)` | Ramp F linearly between limit and target depletion |
| `hcr_ccamlr_krill(gamma, target_depletion, limit_depletion)` | CCAMLR-style precautionary rule |

The hockey-stick rule is particularly useful for CCAMLR-style management:

$$
F =
\begin{cases}
0 & \text{if } B/K \leq b_{\text{limit}} \\
F_{\text{target}} \cdot \frac{B/K - b_{\text{limit}}}{b_{\text{target}} - b_{\text{limit}}} & \text{if } b_{\text{limit}} < B/K < b_{\text{target}} \\
F_{\text{target}} & \text{if } B/K \geq b_{\text{target}}
\end{cases}
$$

## Multi-Area Spatial Models

The framework supports spatial structure in the operating model. Areas are
connected via a movement kernel built from a distance matrix:

```{r spatial-om}
om_spatial <- om_config(
  n_areas         = 2L,
  movement_rate   = 0.1,
  movement_cost_matrix = matrix(c(0, 100, 100, 0), nrow = 2),
  true_params     = list(
    r = 0.3, K = 5000, m = 2,
    sigma_obs = 0.2, q = 0.001,
    B_initial = c(2500, 2500) # per-area initial biomass (totals K)
  )
)
```

The movement kernel is computed from the distance matrix:

```{r movement-kernel}
kernel <- build_movement_kernel(
  movement_cost_matrix = matrix(c(0, 100, 100, 0), nrow = 2),
  decay = 0.01
)
print(kernel)
```

### Spatial Robustness: Self-Test

A *self-test* verifies that when the EM has the same spatial structure as the
OM, the management procedure performs as expected. This is a sanity check
before testing spatial misspecification:

```{r self-test, eval = FALSE}
st <- run_self_test(
  operating_model      = om,
  harvest_control_rule = hcr_const,
  n_sims               = 20L,
  n_proj_years         = 20L,
  seed                 = 123L
)
print(st)
```

If `biomass_conserved` and `positive_biomass` are both `TRUE`, the model
machinery is working correctly.

## Implementation Error

The `impl_error()` specification adds realism by modelling the difference
between the intended TAC and realised catch:

```{r impl-error-example}
ie_ar1 <- impl_error(
  cv          = 0.15,     # coefficient of variation
  bias        = 1.0,      # no systematic bias
  autocorr    = 0.3,      # year-to-year correlation
  max_overage = 1.2       # cap realised catch at 120% of TAC
)

# Apply to a single TAC
set.seed(1)
result_ie <- apply_implementation_error(
  tac          = 200,
  impl_config  = ie_ar1,
  previous_eps = 0,
  seed         = 1L
)
cat(sprintf("TAC = 200, Realised catch = %.1f\n", result_ie$catch))
```

## Observation Error

Observation error parameters can be calibrated from a previously fitted
model, or specified directly:

```{r obs-error-manual}
obs_params <- structure(
  list(
    sigma  = 0.2,
    rho    = 0,
    labels = NULL,
    n_obs  = 20L
  ),
  class = "obs_error_params"
)
print(obs_params)
```

When a fitted `ProductionModel` is available, use `calibrate_observation_error()`
to extract residual-based parameters automatically.

## Model Formulation

The operating model uses the Pella-Tomlinson surplus production model:

**Production function (standard Pella-Tomlinson):**

$$P(B) = \frac{r \, B}{m - 1} \left[ 1 - \left(\frac{B}{K}\right)^{m-1} \right]$$

where *r* is the intrinsic growth rate (Schaefer form $r\,B\,(1 - B/K)$ at
$m = 2$; Fox form $r\,B\,\log(K/B)$ as $m \to 1$).

**State equation:**

$$B_{t+1} = B_t + P(B_t) - C_t$$

**For multi-area models**, biomass is redistributed by a row-stochastic
movement kernel $\mathbf{K}$; arrivals use its transpose so total biomass is
conserved:

$$\mathbf{B}'_{t+1} = (1 - \rho)\,\mathbf{B}_{t+1} + \rho\,\mathbf{K}^\top \mathbf{B}_{t+1}$$

**Reference points** (standard Pella-Tomlinson):

$$
B_{\text{MSY}} = K \left(\frac{1}{m}\right)^{1/(m-1)}, \quad
F_{\text{MSY}} = \frac{r}{m}, \quad
\text{MSY} = F_{\text{MSY}} \times B_{\text{MSY}}
$$

These are computed by `SurplusProductionModel::pella_tomlinson_reference_points()`,
shared with the operating model, estimation model, and performance metrics so
that all components use one consistent set of reference points.

## Next Steps

- Add an estimation model via `em_config()` to test assessment-based feedback
- Explore spatial misspecification (e.g. 2-area OM with 1-area EM)
- Run `compare_scenarios()` to evaluate Pareto-optimal management strategies
- Use `equilibrium_f()` to compute equilibrium fishing mortality at any
  depletion level
